HyperLogLog: Exponentially Bad in Adversarial Settings

### 2020-04-24 docs/HLL_exp_bad/main.tex
work on section 2

\title{\huge HyperLogLog: Exponentially Bad in Adversarial Settings}
\author{Mathilde Raynal, Kenneth G. Paterson}

@@

\section{Introduction}
Naive approaches to computing the cardinality of a big data set, such as sorting the elements or simply maintaining the set of unique elements seen, are impractical at scale. The HyperLogLog algorithm (HLL henceforth)~\cite{HLL } is today the most widely used cardinality estimator. It is increasingly used in settings where adversaries may have  incentives to manipulate the estimate made by the algorithm. For example,~\cite{} proposes its use for detecting network scanning attacks, while Facebook engineers have reported~\cite{} that they use HLL for estimating the number of active users. These applications demand a careful evaluation of the performance of HLL under adversarial input.

@@

\kenny{M: Please fill in missing references above.}

The rest of the paper is organized as follows. Section~\ref{sec:overview} gives an overview of the HyperLogLog algorithm. Section~\ref{sec:attacks} presents our different adversarial models. There, we also give attacks and evaluate their impacts under the different adversarial models. Section~\ref{sec:conclusions} provides our conclusions and some ideas for future work.

\section{HyperLogLog}\label{sec:overview}
\subsection{Algorithm}
HLL is a streaming algorithm based on the bit pattern observation that for a stream of randomly distributed numbers, if we observe a number $x$ with the maximum of $k$ leading 0 bits, the cardinality of the stream is likely to be 2$^{k+1}$. In practice, to guaranty uniform distribution of the inputs, they go through a hash function $h$ before being processed. This simple approach has the disadvantage of a large variance. In order to improve the estimate, we can use many estimators (buckets) instead of one and average the results. To map a value $h(x)$ to a bucket, \cite{loglog} suggests using the first few $n$ bits of $h(x)$ as an index, and then compute the longest sequence of leading 0s on the $l = len(h(x)) - n $ bits left. Estimators from all buckets are averaged with an harmonic mean and scaled by a constant $\alpha$, empirically computed to correct a systematic multiplicative bias. As result, the HyperLogLog algorithm is able to estimate cardinalities greater than $10^9$ with a typical standard error of 2\%, using 1.5 kB of memory.

An interesting property of the HyperLogLog sketch is its ability to be merged in a lossless way. To combine the two buckets at index $i$, take the maximum estimator of the two and assign that to the matching bucket $i$ in the unioned HLL. Such a simple set union operation allows us to easily parallelize operations among multiple machines independently using the same hash function and the same number of buckets.

\subsection{Applications}
For completeness and motivation, we provide examples of concrete uses of HyperLogLog:
\begin{itemize}
    \item Cardinality estimation provides a significantly faster way for Facebook to compute statistics on their users. They use HyperLogLog to find out how many distinct people visited their website in the past week \cite{fbhll}.
    \item A solution to detecting DOS attacks using a sliding HyperLogLog \cite{slidinghll} (a variation which estimates the cardinality over a moving time window) is presented in \cite{portscanhll}. The sketch is used to count how many different ports are used as destination over all packets received, as an attempt to identify when a port scan attack is happening.
    \item \cite{flexswitch} implements the HyperLogLog on a light FlexSwitch to approximate the number of distinct packets in traffic flows and overcome the switch resource limits. The number of unique flows traversing a switch can then be used to provide a better congestion control protocol.
\end{itemize}

Regardless of the sensitivity of the applications, none of them are clear about their threat model, especially for the HyperLogLog. By not assessing the risks involved with the presence of an adversary targeting the HLL sketch, they may allow an attack like the one we present, breaking down the whole solution.

\subsection{Related Work}
The work of \cite{cardestprivacy} breaks the privacy of the HLL sketch and concludes that having the ability to test the insertion of a targeted item gives significant confidence about whether the item was originally present in the input data set. HLL sketches can be used by organizations to store and process location data, which is inherently sensitive, so loosing privacy would leak one user's location and potentially cause great harm. On the other hand, \cite{hllvuln} exploits a vulnerability of the HLL algorithm to manipulate the cardinality estimate, leading to a five-fold reduction from the true cardinality.

Both propose mitigations using a salted sketch, either as replacement, but at the cost of loosing mergeability, or in addition to an unsalted sketch, inducing a memory overhead. This might not be a convenient solution for some applications, they thus encourage further research.

>>>>

\title{HyperLogLog: Exponentially Bad in Adversarial Settings}
\author{Mathilde Raynal, Kenneth G. Paterson}

@@

\section{Introduction}
Naive approaches to computing the cardinality of a big data set, such as sorting the elements or simply maintaining the set of unique elements seen, are impractical at scale. The HyperLogLog algorithm (HLL henceforth)~\cite{loglog,hll2} is today the most widely used cardinality estimator. It is increasingly used in settings where adversaries may have  incentives to manipulate the estimate made by the algorithm. For example,~\cite{portscanhll} proposes its use for detecting network scanning attacks, while Facebook engineers have reported~\cite{fbhll} that they use HLL for estimating the number of active users. These applications demand a careful evaluation of the performance of HLL under adversarial input.

@@

The rest of the paper is organised as follows. Section~\ref{sec:overview} gives an overview of the HyperLogLog algorithm. Section~\ref{sec:attacks} presents our different adversarial models. There, we also give attacks and evaluate their impacts under the different adversarial models. Section~\ref{sec:conclusions} provides our conclusions and some ideas for future work.

\section{Overview of HLL}\label{sec:overview}
\subsection{The Algorithm}
HLL~\cite{loglog,hll2} is a streaming algorithm based on the key observation that, for a stream of randomly distributed values represented as bit-strings of some fixed length, if we observe a value with the maximum of $k$ leading zero-bits, then the cardinality of the stream (i.e.\ the number of distinct values it contains) is likely to be on the order of $2^{k+1}$. In practice, to ensure a uniform distribution, the stream values are first processed by a hash function $h$ before the leading bits are inspected. So to estimate the cardinality of a stream, we need only store a representation of $k$, the length of the largest observed string of leading zero-bits for numbers in the stream. This provides a very compact mechanism for estimating the stream cardinality -- if the true cardinality is $N$, then just $\log\log(N)$ bits are needed to store the estimate.

This simple approach suffers from large variance in the estimation. In order to improve the estimate, we can use many estimators in parallel instead of one, and average the results. Each estimate is stored in its own bucket. To map a value $x$ to one of $2^n$ buckets, \cite{loglog} suggests using the first $n$ bits of $h(x)$ as a bucket index, and then computing the longest sequence of leading zero bits on the remaining $\ell$ bits of $h(x)$ (so that the output length of $h(\cdot)$ is $n+\ell$). Estimators from all buckets are averaged using the harmonic mean and scaled by a constant $\alpha$, where $\alpha$ empirically computed to correct a systematic multiplicative bias. As result, HLL is able to estimate cardinalities greater than $10^9$ with a typical standard error of 2\%, using only 1.5 kB of memory~\cite{add-ref-for-this-claim}.

An interesting property of HLL is that it supports merging in a lossless way. To combine two buckets at index $i$ in two different HLL instances, take the maximum of the two bucket entries and assign that to the matching bucket $i$ in the merged HLL. Such a simple set union operation allows easy parallelisation operations among multiple machines independently, provided they use the same hash function and the same number of buckets.

\subsection{Applications}
For completeness and motivation, we provide examples of concrete uses of HLL:
\begin{itemize}
    \item Cardinality estimation provides a significantly faster way for Facebook to compute statistics on their users. They use HLL to find out how many distinct people visited their website in the past week~\cite{fbhll}.
    \item HLL has been proposed as a solution to detecting Denial-of-Service attacks~\cite{portscanhll} (this paper actually uses a variation on HLL called sliding HLL~\cite{slidinghll} which estimates the cardinality over a moving time window). Specifically, HLL is used to count how many different ports are used as destination port over all packets received, as an attempt to identify when a port scan attack is underway.
        \item HLL has been implemented in network switches to approximate the number of distinct packets in traffic flows and overcome the switch resource limits~\cite{flexswitch}. The number of unique flows traversing a switch can then be used to provide a better congestion control protocol.
        \end{itemize}

Regardless of the sensitivity of the above examples, none of the relevant papers are clear about their threat model, especially for the HLL component.  They are all potentially vulnerable to the attacks we present below.
%By not assessing the risks involved with the presence of an adversary targeting the HLL sketch, they may allow an attack like the one we present, breaking down the whole solution.

\subsection{Related Work}

The work of~\cite{cardestprivacy} breaks the privacy of the HLL sketch and concludes that having the ability to test the insertion \kenny{what does this mean?} of a targeted item gives information about whether the item was originally present in the input data set. It is argued in~\cite{cardestprivacy} that HLL can be used by organizations to store and process location data, which is inherently sensitive, so losing privacy would leak users' locations and potentially cause great harm.

In~\cite{hllvuln}, Reviriego and Ting exploit a vulnerability of HLL to manipulate the cardinality estimate, leading to a five-fold reduction in the estimate compared to the true cardinality.

Both~\cite{cardestprivacy,hllvuln} propose mitigations using a salted sketch, either as replacement (but at the cost of loosing mergability) or in addition to an unsalted sketch (inducing a memory overhead). 
%This might not be a convenient solution for some applications, they thus encourage further research.
